---
title: "Classifying soil texture with classify_texture()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classifying soil texture with classify_texture()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r packages, message=FALSE}
library(tidysoiltexture)
library(dplyr)
library(tibble)
```

## Overview

`classify_texture()` assigns USDA texture classes (sand, loam, clay, etc.) to
soil samples based on their sand, silt, and clay fractions. It is an S3
generic with three dispatch methods:

| Input type | What it dispatches to |
|------------|----------------------|
| Data frame / tibble | `classify_texture.default()` |
| `sf` point object | `classify_texture.sf()` |
| `terra` SpatRaster | `classify_texture.SpatRaster()` |

All methods accept and return tibbles (or the spatial equivalent), and all
support tidy evaluation — you can pass bare column names or string names
interchangeably.

---

## 1. Basic usage

```{r basic}
soils <- tribble(
  ~sample_id, ~sand, ~silt, ~clay,
  "A1",          65,    20,    15,
  "A2",          20,    50,    30,
  "A3",           8,    72,    20,
  "A4",          40,    38,    22,
  "A5",          52,     6,    42,
  "A6",           6,    47,    47
)

classify_texture(soils, sand, silt, clay)
```

Two columns are appended to the input tibble:

* `.texture_class` — full USDA class name
* `.texture_abbr` — standard abbreviation (e.g. `SaLo`, `Cl`)

---

## 2. Tidy evaluation

Column names can be passed as bare symbols or as string-named columns.
The function uses `rlang::enquo()` internally so it integrates naturally
into dplyr pipelines.

```{r tidyeval-pipeline}
# Works inside a pipe with renamed columns
soils |>
  rename(s = sand, si = silt, c = clay) |>
  classify_texture(sand = s, silt = si, clay = c)
```

Scalar values also work for any argument — useful for constant fractions
across a dataset:

```{r scalar-args}
# All samples have the same silt (25%); sand and clay vary
tibble(sand = c(60, 40, 15), clay = c(15, 35, 60)) |>
  classify_texture(sand = sand, silt = 25, clay = clay)
```

---

## 3. Validation

`classify_texture()` validates that the three fractions sum to 100 (within
a tolerance of ± 0.5). Rows that don't sum correctly raise an error.

```{r validation, error=TRUE}
# Bad row: fractions sum to 95
tibble(sand = 50, silt = 30, clay = 15) |>
  classify_texture(sand, silt, clay)
```

---

## 4. The 12 USDA texture classes

`usda_texture_classes` contains the polygon boundary data used internally.

```{r usda-data}
usda_texture_classes |>
  distinct(class, abbr) |>
  arrange(class)
```

A helper lookup table of the 12 classes with representative centroids:

```{r centroids}
tribble(
  ~Class,            ~Abbr,     ~Sand, ~Silt, ~Clay,
  "sand",            "Sa",         92,     5,     3,
  "loamy sand",      "LoSa",       80,    12,     8,
  "sandy loam",      "SaLo",       65,    22,    13,
  "loam",            "Lo",         41,    39,    20,
  "silty loam",      "SiLo",       20,    65,    15,
  "silt",            "Si",          7,    85,     8,
  "sandy clay loam", "SaClLo",     60,    13,    27,
  "clay loam",       "ClLo",       32,    34,    34,
  "silty clay loam", "SiClLo",     10,    56,    34,
  "sandy clay",      "SaCl",       52,     6,    42,
  "silty clay",      "SiCl",        6,    47,    47,
  "clay",            "Cl",         20,    20,    60
) |>
  classify_texture(Sand, Silt, Clay) |>
  select(Class, Abbr, .texture_class, .texture_abbr)
```

All 12 centroids classify correctly — the `.texture_class` column matches
the `Class` column.

---

## 5. Batch classification and performance

The vectorised backend processes all N points per class simultaneously,
not row by row. This makes it practical for large datasets.

```{r performance}
set.seed(5519)
big <- tibble(
  sand = runif(10000, 5, 85),
  clay = runif(10000, 5, 45)
) |>
  mutate(silt = 100 - sand - clay) |>
  filter(silt >= 2) |>
  slice_head(n = 10000)

elapsed <- system.time(
  result <- classify_texture(big, sand, silt, clay)
)["elapsed"]

cat(sprintf("10 000 samples in %.3f s\n", elapsed))
cat(sprintf("NAs: %d\n", sum(is.na(result$.texture_class))))
```

---

## 6. sf dispatch

If your data is an `sf` point object, `classify_texture()` detects the class
and dispatches to the `.sf` method. Geometry is preserved.

```{r sf-dispatch, eval=requireNamespace("sf", quietly=TRUE)}
library(sf)

pts_sf <- st_as_sf(
  soils,
  coords = c("sand", "silt"),   # using sand/silt as dummy coordinates
  crs    = NA_crs_
)
pts_sf$sand <- soils$sand
pts_sf$silt <- soils$silt
pts_sf$clay <- soils$clay

out_sf <- classify_texture(pts_sf, sand = sand, silt = silt, clay = clay)
class(out_sf)         # still an sf object
names(out_sf)         # .texture_class and .texture_abbr appended
```

```{r sf-skip, eval=!requireNamespace("sf", quietly=TRUE), echo=FALSE}
message("sf not installed — skipping Section 6.")
```

---

## 7. SpatRaster dispatch

For raster stacks (e.g. from SoilGrids), `classify_texture()` dispatches to
the `.SpatRaster` method and returns a categorical raster with 12 levels.

```{r spatraster-dispatch, eval=requireNamespace("terra", quietly=TRUE)}
library(terra)

# Build a small 10×10 test raster
r <- rast(ncols = 10, nrows = 10, nlyrs = 3,
          xmin = 0, xmax = 1, ymin = 0, ymax = 1)
names(r) <- c("sand", "silt", "clay")

set.seed(7712)
sand_v <- runif(100, 10, 70)
clay_v <- runif(100, 5, 40)
silt_v <- 100 - sand_v - clay_v
values(r) <- cbind(sand_v, silt_v, clay_v)

r_class <- classify_texture(r, sand = "sand", silt = "silt", clay = "clay")
print(r_class)
levels(r_class)[[1]]
```

```{r spatraster-skip, eval=!requireNamespace("terra", quietly=TRUE), echo=FALSE}
message("terra not installed — skipping Section 7.")
```

**SoilGrids note:** SoilGrids stores fractions in g/kg (0–1000). Divide by 10
before classifying. `classify_texture()` will emit a `cli_warn()` if any value
exceeds 100, so the error surfaces clearly.

```r
r_sg <- rast("SoilGrids_sand_silt_clay.tif")   # g/kg layers
r_class <- classify_texture(r_sg / 10,
                             sand = "sand", silt = "silt", clay = "clay")
```

---

## 8. Edge cases: boundary points and vertices

The ray-casting algorithm includes a collinearity check so points that fall
exactly on class boundaries (shared edges, polygon vertices) are classified
correctly rather than returning `NA`.

```{r boundary}
# Clay/silty clay boundary at clay=40%, silt=60%
boundary_pts <- tribble(
  ~description,          ~sand, ~silt, ~clay,
  "clay vertex",             0,    40,    60,
  "silt/silty clay border",  0,    60,    40,
  "clay loam centroid",     32,    34,    34
)

classify_texture(boundary_pts, sand, silt, clay) |>
  select(description, .texture_class)
```

---

## See also

* `vignette("gg-texture-triangle")` — visualise classified samples on a
  texture triangle
* `vignette("gis-workflow")` — full GIS workflow with sf and terra
