---
title: "GIS Workflow: Classifying Raster and Vector Soil Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GIS Workflow: Classifying Raster and Vector Soil Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 5,
  dpi = 120,
  out.width = "80%",
  fig.align = "center"
)
have_terra <- requireNamespace("terra", quietly = TRUE)
have_sf    <- requireNamespace("sf",    quietly = TRUE)
```

As R-GIS workflows become more common in soil science, `tidysoiltexture`
provides `classify_texture()` methods for two key spatial classes:

* **`terra::SpatRaster`** — classify every cell of a sand/silt/clay raster
  stack and return a categorical output raster suitable for mapping and export.
* **`sf`** — classify point features (e.g. field observations, pedon
  locations) and return the same `sf` object with texture class columns
  appended.

```{r prereqs}
library(tidysoiltexture)
```

> **`terra`** and **`sf`** are suggested dependencies. Install them with
> `install.packages(c("terra", "sf"))` to follow the full workflow.

---

## 1. Raster workflow with `terra`

### 1.1 Load a sand / silt / clay raster stack

In practice your raster stack typically comes from a downloaded product such
as [SoilGrids](https://soilgrids.org) or
[ESDAC](https://esdac.jrc.ec.europa.eu). The only requirement is that the
`SpatRaster` has named layers for sand, silt, and clay (in %).

```{r load-real, eval = FALSE}
library(terra)

# Load a 3-band GeoTIFF where bands are named sand, silt, clay
r <- terra::rast("path/to/soil_texture_stack.tif")

# Rename layers if needed
names(r) <- c("sand", "silt", "clay")

# Inspect
print(r)
terra::plot(r, main = c("Sand (%)", "Silt (%)", "Clay (%)"))
```

For this vignette we build a synthetic stack that mimics a 20 × 20 km
catchment with a sandy upland area and a clay-rich valley floor:

```{r make-demo, eval = have_terra}
library(terra)

set.seed(1847)

# 50 × 50 cell grid, 400 m resolution, EPSG:32632 (UTM zone 32N)
r_template <- terra::rast(
  nrows = 50, ncols = 50,
  xmin = 0, xmax = 20000,
  ymin = 0, ymax = 20000,
  crs  = "EPSG:32632"
)

# Simulate a north–south valley: clay increases toward centre column
col_idx  <- terra::colFromX(r_template, terra::xFromCol(r_template,
                              seq_len(terra::ncol(r_template))))
valley   <- dnorm(seq(0, 1, length.out = terra::ncol(r_template)),
                  mean = 0.5, sd = 0.18)
valley_m <- matrix(rep(valley, each = terra::nrow(r_template)),
                   nrow = terra::nrow(r_template))

# Sand decreases toward valley; clay increases
sand_m <- pmax(15, pmin(85, 70 - 45 * valley_m +
               matrix(rnorm(2500, 0, 5), 50, 50)))
clay_m <- pmax(5,  pmin(55, 10 + 40 * valley_m +
               matrix(rnorm(2500, 0, 4), 50, 50)))
silt_m <- 100 - sand_m - clay_m

# Force silt ≥ 2 (push excess into sand)
excess <- pmin(0, silt_m - 2)
sand_m <- sand_m + excess
silt_m <- silt_m - excess

# Assemble into SpatRaster
r <- c(terra::setValues(r_template, sand_m),
       terra::setValues(r_template, silt_m),
       terra::setValues(r_template, clay_m))
names(r) <- c("sand", "silt", "clay")

print(r)
```

### 1.2 Inspect the stack

```{r inspect-stack, eval = have_terra}
terra::summary(r)
terra::plot(r,
  main      = c("Sand (%)", "Silt (%)", "Clay (%)"),
  col       = grDevices::hcl.colors(50, "YlOrBr", rev = TRUE),
  axes      = FALSE
)
```

### 1.3 Classify

Pass the `SpatRaster` directly to `classify_texture()`. The `sand`, `silt`,
and `clay` arguments name the layers (default `"sand"`, `"silt"`, `"clay"`
so they can be omitted here):

```{r classify-raster, eval = have_terra}
r_class <- classify_texture(r)   # layer names match defaults
print(r_class)
```

The output is a single-layer categorical `SpatRaster`. Integer cell values
are mapped to texture class names via a raster attribute table (RAT):

```{r inspect-levels, eval = have_terra}
terra::levels(r_class)[[1]]
```

Count how many cells fall into each class:

```{r freq-table, eval = have_terra}
terra::freq(r_class) |>
  dplyr::arrange(dplyr::desc(count))
```

### 1.4 Plot the classified raster

`terra::plot()` recognises categorical rasters and automatically draws a
colour legend:

```{r plot-class, eval = have_terra}
terra::plot(r_class,
  main  = "USDA Texture Classification",
  axes  = FALSE,
  mar   = c(2, 1, 2, 8)   # extra right margin for legend
)
```

Side-by-side with the sand layer to confirm the spatial pattern:

```{r plot-comparison, eval = have_terra, fig.width = 9, fig.height = 4.5, out.width = "100%"}
oldpar <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 6))

terra::plot(r[["sand"]],
  main = "Sand (%)",
  col  = grDevices::hcl.colors(50, "YlOrBr", rev = TRUE),
  axes = FALSE)

terra::plot(r_class,
  main = "Texture class",
  axes = FALSE,
  mar  = c(1, 1, 2, 10))

par(oldpar)
```

### 1.5 Export the classified raster

The output is a standard `SpatRaster` and can be written to any
GDAL-supported format. For GeoTIFF:

```{r export, eval = FALSE}
terra::writeRaster(r_class, "texture_classification.tif",
                   datatype = "INT1U", overwrite = TRUE)
```

---

## 2. Vector workflow with `sf`

For point observations (e.g. soil survey pedons, lab samples with
coordinates) the `sf` method appends `.texture_class` and `.texture_abbr`
while preserving geometry:

```{r sf-classify, eval = have_sf}
library(sf)

# 30 synthetic pedon locations across the study area
set.seed(3312)
pedons <- sf::st_as_sf(
  tibble::tibble(
    pedon_id = sprintf("P%02d", 1:30),
    x        = runif(30, 500,  19500),
    y        = runif(30, 500,  19500),
    sand     = runif(30, 10,   80),
    clay     = runif(30,  5,   45)
  ) |>
    dplyr::mutate(silt = 100 - sand - clay) |>
    dplyr::filter(silt >= 2),
  coords = c("x", "y"),
  crs    = "EPSG:32632"
)

# Classify — bare column names, tidy-eval, geometry preserved
pedons_class <- classify_texture(pedons, sand = sand, silt = silt, clay = clay)

dplyr::select(pedons_class, pedon_id, sand, silt, clay,
              .texture_class, .texture_abbr)
```

### 2.1 Plot points over the classified raster

```{r sf-overlay, eval = have_terra && have_sf, fig.width = 6, fig.height = 5.5}
terra::plot(r_class,
  main = "Texture class + pedon locations",
  axes = FALSE,
  mar  = c(1, 1, 2, 10)
)
plot(sf::st_geometry(pedons_class),
  add  = TRUE,
  pch  = 21,
  bg   = "white",
  cex  = 0.9
)
```

---

## 3. Notes on real-world data sources

| Source | Resolution | Coverage | URL |
|---|---|---|---|
| **SoilGrids 2.0** | 250 m | Global | soilgrids.org |
| **ESDAC LUCAS** | Point / 500 m | Europe | esdac.jrc.ec.europa.eu |
| **SSURGO** | Polygon | USA | websoilsurvey.nrcs.usda.gov |

SoilGrids delivers sand/silt/clay as separate GeoTIFFs (e.g.
`sand_0-5cm_mean.tif`). After downloading, stack and rename:

```{r soilgrids-example, eval = FALSE}
library(terra)

r <- terra::rast(c(
  "sand_0-5cm_mean.tif",
  "silt_0-5cm_mean.tif",
  "clay_0-5cm_mean.tif"
))
names(r) <- c("sand", "silt", "clay")

# SoilGrids values are in g/kg — convert to %
r <- r / 10

# Classify
r_class <- classify_texture(r)
terra::writeRaster(r_class, "usda_texture_class.tif",
                   datatype = "INT1U", overwrite = TRUE)
```
